Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit

### 2021-09-14 submission/manuscript.rmd
replace moderate linear interaction with absence of treatment effect in figures

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In 
 each sample we predicted absolute benefit using: models with the PI and a 
 constant relative treatment effect; models including an interaction of treatment 
 with the PI; stratification in quarters of the PI; nonlinear transformations of 
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach 
 using Akaike’s Information Criterion. We evaluated predictive performance using 
 root mean squared error and measures of discrimination and calibration for 
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds 
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect 
 strength, and the PI's discriminative ability. **Results:** Models including a 
 PI by treatment interaction performed better under most simulation settings. 
 Flexible models required larger sample sizes and higher AUC of the PI to 
 outperform the linear interaction model. The adaptive approach performed 
 similarly to the best-performing method in most scenarios. **Conclusion:** 
 Under most circumstances, a model with a linear interaction of the PI with 
 treatment is the optimal risk-based approach to predicting absolute treatment
 benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The adaptive approach had adequate performance, following closely the
performance of the "true" model in most scenarios. However, with smaller sample
sizes it tended to "miss" the treatment-risk interactions and selected simpler
models. This resulted in increased RMSE variability in these scenarios,
especially in the case of true strong linear deviations from the base case
scenario. This tendency was mitigated when the discriminative ability of the
prediction model was higher. Therefore, in the case of smaller sample sizes
and/or poorly discriminating prediction models, the simpler linear interaction
model is a safer choice for predicting absolute benefits.

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the PI's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The adaptive approach had adequate performance, following closely the
performance of the "true" model in most scenarios. However, with smaller sample
sizes it tended to "miss" the treatment-risk interactions and selected simpler
models (Supplementary Table S7). This resulted in increased RMSE variability in
these scenarios, especially in the case of true strong linear deviations from
the base case scenario. This tendency was mitigated when the discriminative
ability of the prediction model was higher. Therefore, in the case of smaller
sample sizes and/or poorly discriminating prediction models, the simpler linear
interaction model is a safer choice for predicting absolute benefits.

### 2021-09-14 submission/manuscript.tex
update title
replace moderate linear interaction with absence of treatment effect in figures

\title{A model including an interaction of treatment with baseline risk
was the optimal risk-based approach to predicting absolute treatment
benefit}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
performed similarly to the best-performing method in most scenarios.
\textbf{Conclusion:} Under most circumstances, a model with a linear
interaction of the PI with treatment is the optimal risk-based approach
to predicting absolute treatment benefit.
\end{abstract}

@@

The adaptive approach had adequate performance, following closely the
performance of the ``true'' model in most scenarios. However, with
smaller sample sizes it tended to ``miss'' the treatment-risk
interactions and selected simpler models. This resulted in increased
RMSE variability in these scenarios, especially in the case of true
strong linear deviations from the base case scenario. This tendency was
mitigated when the discriminative ability of the prediction model was
higher. Therefore, in the case of smaller sample sizes and/or poorly
discriminating prediction models, the simpler linear interaction model
is a safer choice for predicting absolute benefits.

>>>>

\title{Linear interaction of treatment with baseline risk was sufficient
for predicting individualized benefit}

@@

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
tended to select simpler models with smaller sample sizes and/or worse
prediction performance. \textbf{Conclusion:} Under most circumstances, a
model with a linear interaction of the PI with treatment is the optimal
risk-based approach to predicting absolute treatment benefit.
\end{abstract}

@@

The adaptive approach had adequate performance, following closely the
performance of the ``true'' model in most scenarios. However, with
smaller sample sizes it tended to ``miss'' the treatment-risk
interactions and selected simpler models (Supplementary Table S7). This
resulted in increased RMSE variability in these scenarios, especially in
the case of true strong linear deviations from the base case scenario.
This tendency was mitigated when the discriminative ability of the
prediction model was higher. Therefore, in the case of smaller sample
sizes and/or poorly discriminating prediction models, the simpler linear
interaction model is a safer choice for predicting absolute benefits.
